---
title: "Analysing MICS 2010 for Bhutan"
subtitle:  "NSB course - Paro"
author: "Christophe Bontemps SIAP"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: tango
    toc: yes
    toc_float: true
    keep_md: no
    number_sections: no
    theme: lumen
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( message = FALSE, warning = FALSE, results =TRUE, echo = TRUE) 

# Fix tabularray issue globally
options(modelsummary_format_numeric_latex = "plain")
options(knitr.table.format = "simple")  

```


### Good practices and file management {-}

We need a minimum of organisation in the data and code folders, as well as some R packages.
The folder is organized  with:

- Data  <- *Original (raw) data will be there*
- Outputs <- *Any result will go there*
- Documentation  <- *Docs and explanations, additional resources*

In our organization, the code (.R and .Rmd files) will be at the root of the directory. In general, it is recommended to have a directory with `code`. 

We also need some packages: 

```{r packages}
# Importing SPSS file
library(haven)

# Tidy data management packages
library(tidyverse)
library(data.table)

# Plotting packages
library(ggplot2)
library(RColorBrewer)

# Nice presentation of results
library(knitr)
library(papeR)
library(visreg)
library(ggpubr)

# Tables 
library(modelsummary)
library(gt) 
library(gtsummary)

```

#  Understanding MICS

There are several files available for [MICS 2010](https://mics.unicef.org/surveys?display=card&keys=Bhutan). Five SPSS data files were produced, corresponding to the units of analysis:

- Households -` hh.sav`
- Household members - `hl.sav`
- Women in reproductive age (15-49 years of age) –`wm.sav`
- Mothers or primary caretakers of children under the age of five – `ch.sav``

> Here we will focus on MICS Women's questionnaire and analyze child marriage

## Reading the data

Our data set being in the directory `Data/MICSBhutan`, we need to specify the right path when loading the data. Here the data are in SPSS format ($.sav$) and we need a specific package  (*haven*) to load the data. 

```{r}
## Load data locally
#wm  <- read_spss(file ="Data/MICSBhutan/wm.sav", user_na = TRUE)


## Online download
urlfile <- "https://raw.githubusercontent.com/XtopheB/DataScience4OS/main/Data/MICSBhutan/wm.sav"
wm <- read_spss(file = urlfile, user_na = TRUE)

# As a data frame
wm <- as.data.frame(wm)
```

It is good to have a look at the data, even if it has  **`r ncol(wm)`** variables. 


## Select the variables that may be relevant 

First, we need to redefine some variable and adjust their right type (numeric, character, logic, etc.). For that, we have to look at the [data dictionary](https://catalog.ihsn.org/catalog/2877/data-dictionary/F5?file_name=wm).



> We create a new data frame with only the variables relevant to our analysis 

```{r}
#select the only relevant variables
df <- select(wm, 
             WB2, WAGEM, region,  welevel, windex5, WAGEM, MA1, HH7, HH1, wmweight,  WM7)
```

Then we need to keep only observations corresponding to our definition: **Women of age 20 to 24**.

```{r}
# Rename and reformat some variables 
df$age <- as.numeric(as.character(df$WB2))

#Select only women ages between 20 and 24 and those completed the survey
df <- df %>%
  filter(WM7 == 1,
         age >= 20, 
         age < 25)

```

That data frame has now **`r nrow(df)`** observations and  **`r ncol(df)`** variables 

### Defining the variables with the right "type" 

R is a ***typed*** language with six primary data *types* that variables can hold (numeric,integer, character,  logical, ...). The type determines what kind of data the variable stores and which operations can be performed on it. 

> It is important to work with a data frame where each variable has the right type

```{r}


# the level of education is also a factor variable with 6 levels
df$welevel <- as.factor(df$welevel)

```


```{r}
# The variables should be of the right type
df <- df %>%
  mutate(wealth = factor(windex5, 
                            levels = c(1, 2, 3, 4, 5),
                            labels = c("Poorest", "Second", "Middle", "Fourth", "Richest")
                             ),
         region = factor(region, 
                            levels = c(1, 2, 3),
                            labels = c("Western", "Central", "Eastern")
                         ), 
         welevel = factor(welevel, 
                           levels = c(1, 2, 3),
                           labels = c("None", "Primary", "Secondary+")) 
           )

# Also creating a binary variable  that records 1 for child marriage and 0 otherwise
# That variable must be a factor variable

df <- df %>%
  mutate(Before18 = if_else(WAGEM < 18, 1, 0),
         Before18 = factor(Before18,
                           levels = c(0, 1),
                           labels = c("Married 18+", "Child Marriage"))
         )


```


```{r}
# Keeping only useful
df <-df %>%
  select(Before18, age, WAGEM, region,  welevel, wealth, WAGEM, MA1, HH7, HH1, wmweight)
```


# Descriptive statistics

```{r}
datasummary_skim(df, type = "numeric")
```

The variable `WAGEM` is the variable that will help us define  child marriage. But it has some missing information, so we should refine the data set. 

```{r}
df <- drop_na(df)
```

> The final data set has **`r nrow(df)`** observations (and  still `r ncol(df)` variables )



```{r}
datasummary_skim(df, type = "categorical")
```

## Cross-tabulations 
We can also do some cross-tabulations to asses the relations between child marriage and education  


```{r}
tab <- table(df$Before18, df$welevel)

# Column percentages
tt <- prop.table(tab, margin = ) * 100  # multiply by 100 for % 
kable(tt, digits =2, caption = "Child marriage vs Education")
```


```{r}
tab <- table(df$Before18, df$region)

# Column percentages
tt <- prop.table(tab, margin = 2) * 100  # multiply by 100 for % 
kable(tt, digits =2, caption = "Child marriage vs Area")
```


# Logistic regression

We will use the DHS survey at the individual level and construct a binary response variable indicating whether there was **marriage before 18 years** old (or "*girl's marriage*"). We will try to understand which factor**S** may affect the probability to observe girl's marriage, such as education, wealth and other socioeconomic factors.   



```{r}
# Defining the variables of the model
Y<-"Before18"               # Response variable


XCovars <- c("age", "welevel", "wealth", "region")
formula_string <- paste(Y, paste(XCovars, collapse = " + "), sep = " ~ ")

print(paste(" Regression formula: ",formula_string))

```

## Results {-}
We use a logit model with several explanatory variables

```{r, results='asis'}
# Logistics Regression
glm.fit <- glm(formula_string,
               data = df,
               family = binomial)

# Nice printing of the results (using paper and knitr packages)
pretty_lm2 <- prettify(summary(glm.fit))
kable(pretty_lm2, digits = 3)

```


### Confusion Matrix {-}

The confusion matrix shows how well the model predicted the outcome. If the model was perfect, we would only have elements on the diagonal of this matrix and 0’s everywhere. 

```{r, results=TRUE }
library("regclass")
cm <- confusion_matrix(glm.fit)
cm
```


```{r, results=TRUE }
TN <- cm[1,1]
FP <- cm[2,1]
FN <- cm[1,2]
TP <- cm[2,2]

# Some indicators 
misclass <- round(100* (FP+FN)/ (TN+TP+FN+FP), 1)

accuracy <- round(100* (TP + TN) /(TN+TP+FN+FP),1)
accuracy
```
In total, the model misclassified ( `r FP`   + `r FN` / `r nrow(df)`) = **`r misclass` %** of the outcome, which yields **an accuracy of `r accuracy` per cent** .

### Visual representation of the logistic model{-} 
We can also visualize the effect of some of the most significant variables in the model. 


```{r}
p_educ <- visreg(glm.fit, "welevel", scale="response", rug=0,  # for rugs =2
                 xlab="Education level",
                 ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p_wealth <- visreg(glm.fit, "wealth", scale="response", rug=0,  # for rugs =2
                   xlab="Wealth index",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p_age <- visreg(glm.fit, "age", scale="response", rug=0,  # for rugs =2
                   xlab="Age",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

p_area <- visreg(glm.fit, "region", scale="response", rug=0,  # for rugs =2
                   xlab="Region",
                   ylab="Prob(Child Marriage)", gg=TRUE) +
  ylim(0,1) +theme_minimal()

## Combining the graphics
figure <- ggarrange( p_educ, p_wealth, p_age,  p_area,
                     #labels = c("Education", "wealth"  "Age"),
                     ncol = 2, nrow = 2)

figure
```


We can see the marginal effects  of significant predictors in the model. 

As the level of education increases, the probability decreases. We can notice a clear drop in the probability between the category 2 "primary" and  3 “Secondary+”. 

Surprisingly, all things being equal, the probability of child marriage does not significantly decrease with the increase of the level of wealth. 

The probability of being married before 18 decreases also with age. This shows a lower prevalence of the practice in the past.

Less significant, there seems to be a small effect of the region "central" where the probability is higher. 

# But the analysis is not really "correct"...
There is a sampling protocol and we should integrate the **weights** for each household. For that, we'll use the package  `survey`

```{r}
# Declaring survey sample design/weights and creating tables 
library(survey)
```

Fortunately, we knowthe variables that define the strata and the weights 

```{r}
wmdesign<-svydesign(id= ~HH1,strata= ~HH7, weights = ~wmweight, nest=TRUE, data=df)
```

 We can redo the cross-tabulation, taking into account the weights and stratas..
 
```{r }
library(RcmdrMisc)
# cross-tabulations (instead of correlation matrix)
colPercents(svytable(~Before18+welevel, wmdesign,round=TRUE))
```


```{r }
colPercents(svytable(~Before18+wealth, wmdesign,round=TRUE))
```

```{r}
# Full model with sampling design included
# We use the same model as previously

wm.sglm1 <- svyglm(formula = formula_string, 
                   design = wmdesign, 
                   family = binomial, 
                   data = df)

summary(wm.sglm1)
```


```{r}
# Visualizing regression

pnew_educ <- visreg(wm.sglm1, "welevel", scale="response", rug=0,  # for rugs =2
                 xlab="Education level",
                 ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

pnew_wealth <- visreg(wm.sglm1, "wealth", scale="response", rug=0,  # for rugs =2
                   xlab="Wealth index",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()

pnew_age <- visreg(wm.sglm1, "age", scale="response", rug=0,  # for rugs =2
                   xlab="Age",
                   ylab="Prob(Child Marriage)", gg=TRUE) + 
  ylim(0,1) +theme_minimal()


pnew_area <- visreg(wm.sglm1, "region", scale="response", rug=0,  # for rugs =2
                   xlab="Region",
                   ylab="Prob(Child Marriage)", gg=TRUE) +
  ylim(0,1) +theme_minimal()

## Combining the graphics
New_figure <- ggarrange( pnew_educ, pnew_wealth, pnew_age, pnew_area,
                     # labels = c("Education", "wealth" , "Age", "Region"),
                     ncol = 2, nrow = 2)

New_figure
```


# Conclusion

The results suggest that: 

-  “*Education*”, 
- “*Age*”, 
and, to a lesser extend, 
- “*region*” 

are important variables in explaining and predicting the outcome of marriage before 18 in Bhutan.

It is intuitive that “Education” empowers women and it significantly reduces the probability of being married before 18. Surprisingly, "Wealth" does not play a role here, while this is something  often observed in such analysis. Further research is needed to provide insights on these findings. 


```{r}
knitr::knit_exit()
```

