---
title: "Exercise with Parallel Coordinate Plots (PCP)"
subtitle: '*An example*'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
  )
```

# Introduction

Parallel Coordinate Plots (PCP) are a powerful visualization technique for exploring multivariate data. They allow us to visualize multiple dimensions simultaneously by representing each variable as a vertical axis, with individual observations shown as connected lines across these axes.



# Load Libraries
Beofre being able to run this code, we should have 3 packages installed:
- `tidyverse` is one of the main package for data science
- `GGally` is a package that propose Parallel Coordinate Plots (PCP)
- `plotly` is use to create interactive graphics

Once installed, we can use (load) the library to use the many fucntions proposed.  

```{r load-libraries}
library(tidyverse)
library(GGally)      # For ggparcoord
library(plotly)      # For interactive plots
```

# Data 

We generate the data for the sake of the exercise

```{r generate-data}
set.seed(123)

# Create dataset 
data <- tibble(
  id = 1:10,
  sex = c(rep("Men", 6), rep("Women", 4)),
  
  # Age: Men older (45-55), Women younger (25-35)
  age = c(
    48, 52, 45, 50, 47, 54,  # Men: 45-55 range
    28, 32, 26, 30           # Women: 25-35 range
  ),
  
  # Education: Men lower (12-14 years), Women higher (16-18 years)
  education = c(
    12, 14, 13, 12, 14, 13,  # Men: high school to some college
    18, 17, 18, 16           # Women: college to grad school
  ),
  
  # Income: Men high (70k-90k), Women moderate (40k-55k)
  income = c(
    85000, 78000, 82000, 90000, 75000, 88000,  # Men: high income
    48000, 52000, 45000, 50000                 # Women: moderate income
  )
)

# Saving for students
write.csv(select(data, c(age, education, income)),
                 "Data/PCPData.csv", row.names = FALSE)

```


# Basic Parallel Coordinate Plot


```{r basic-pcp}
# Prepare data for PCP (select only numeric variables)
data_pcp <- data %>%
  select(id, age, education, income)

# Basic PCP using ggparcoord
ggparcoord(
  data_pcp,
  columns = 2:4,  # age, education, income
  groupColumn = "id",
  scale = "uniminmax"
) +
  labs(
    title = "Parallel Coordinate Plot: Age, Education, and Income",
    subtitle = "10 observations showing two distinct patterns",
    x = NULL,
    y = "Scaled Values (0-1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50", size = 11)
  )
```

**Observation:** Notice how the lines form two distinct groups, even though we haven't colored them.

# Enhanced PCP with Transparency

Adding transparency (alpha) helps reveal overlapping patterns and density.

```{r enhanced-pcp}
ggparcoord(
  data_pcp,
  columns = 2:4,
  groupColumn = "id",
  scale = "uniminmax",
  alphaLines = 0.6
) +
  geom_line(size = 1.2) +
  labs(
    title = "Enhanced Parallel Coordinate Plot",
    subtitle = "With transparency to show overlapping patterns",
    x = NULL,
    y = "Scaled Values"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    panel.grid.major.x = element_line(color = "grey80", size = 0.5),
    panel.grid.minor = element_blank()
  )
```

# Interactive version


```{r interactive-pcpsex, fig.height=7}
# Prepare data for plotly
plot_data <- data %>%
  mutate(
    color_code = ifelse(sex == "Men", 0, 1),
    label = paste0(
      "ID: ", id, "<br>",
      "Sex: ", sex, "<br>",
      "Age: ", age, "<br>",
      "Education: ", education, " years<br>",
      "Income: $", format(income, big.mark = ",")
    )
  )

n <- nrow(plot_data)

# Create interactive parallel coordinates plot
fig <- plot_ly(
  data = plot_data,
  type = 'parcoords',
  line = list(
    color = rep(0, n),
    colorscale = list(c(0, '#999999'), c(1, '#999999')),
    showscale = FALSE,
    width = 4
  ),
  dimensions = list(
    list(label = 'Age', values = ~age, range = c(20, 60)),
    list(label = 'Education (years)', values = ~education, range = c(10, 20)),
    list(label = 'Income ($)', values = ~income, range = c(30000, 90000))
  )
)

fig <- fig %>%
  layout(
    title = list(
      text = "Interactive Parallel Coordinate Plot",
      font = list(size = 16),
      y = 0.98
    ),
    annotations = list(
      list(
        text = "Drag axes to filter",
        showarrow = FALSE,
        xref = "paper", yref = "paper",
        x = 1, y = 1.2,
        font = list(size = 9, color = "grey40")
      )
    ),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0.5, xanchor = "center",
        y = 1.22, yanchor = "top",
        buttons = list(
          list(
            method = "restyle",
            args = list(list(
              "line.color" = list(rep(0, n)),
              "line.colorscale" = list(list(c(0, '#999999'), c(1, '#999999')))
            )),
            label = "No Color"
          ),
          list(
            method = "restyle",
            args = list(list(
              "line.color" = list(plot_data$color_code),
              "line.colorscale" = list(list(c(0, '#3498DB'), c(1, '#E74C3C')))
            )),
            label = "Color"
          )
        )
      )
    ),
    margin = list(l = 100, r = 100, t = 180, b = 50),
    plot_bgcolor = '#f8f9fa',
    paper_bgcolor = 'white'
  )

fig
```




# PCP with Hidden Grouping

We can color the lines by sex (a variable that was hidden before), making the two groups visually distinct.

```{r pcp-colored,, eval=FALSE, include=FALSE}
# Add sex back for coloring
data_with_sex <- data %>%
  select(id, sex, age, education, income)

ggparcoord(
  data_with_sex,
  columns = 3:5,  # age, education, income
  groupColumn = "sex",
  scale = "uniminmax",
  alphaLines = 0.7
) +
  scale_color_manual(values = c("Men" = "#3498DB", "Women" = "#E74C3C")) +
  labs(
    title = "Parallel Coordinate Plot: Two Distinct Groups",
    subtitle = "Blue and red lines show different patterns (sex not labeled)",
    x = NULL,
    y = "Scaled Values (0-1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 11),
    panel.grid.major.x = element_line(color = "grey80", size = 0.5)
  )
```

**Key Patterns:**

- **Blue lines** (Men): Higher on age and income, moderate on education
- **Red lines** (Women): Lower on age and income, higher on education


# Interactive Parallel Coordinate Plot

Finally, let's create an **interactive version** using `plotly` that allows hovering and filtering.

## Interactive Features

The interactive plot below allows you to:

1. **Hover** over lines to see individual observation details
2. **Click and drag** on any axis to filter the data
3. **Reorder axes** by dragging axis labels
4. **Zoom** into specific ranges

```{r interactive-pcp, fig.height=7}
# Prepare data for plotly
plot_data <- data %>%
  mutate(
    color_code = ifelse(sex == "Men", 0, 1),
    label = paste0(
      "ID: ", id, "<br>",
      "Sex: ", sex, "<br>",
      "Age: ", age, "<br>",
      "Education: ", education, " years<br>",
      "Income: $", format(income, big.mark = ",")
    )
  )

# Create interactive parallel coordinates plot
fig <- plot_ly(
  data = plot_data,
  type = 'parcoords',
  line = list(
    color = ~color_code,
    colorscale = list(
      c(0, '#3498DB'),  # Men: Blue
      c(1, '#E74C3C')   # Women: Red
    ),
    showscale = FALSE
  ),
  dimensions = list(
    list(
      label = 'Age',
      values = ~age,
      range = c(20, 60)
    ),
    list(
      label = 'Education (years)',
      values = ~education,
      range = c(10, 20)
    ),
    list(
      label = 'Income ($)',
      values = ~income,
      range = c(30000, 90000)
    )
  )
)

fig <- fig %>%
  layout(
    title = list(
      text = "Interactive Parallel Coordinate Plot<br><sub>Hover over lines to see details | Drag axes to filter</sub>",
      font = list(size = 16)
    ),
    margin = list(l = 100, r = 100, t = 100, b = 50),
    plot_bgcolor = '#f8f9fa',
    paper_bgcolor = 'white'
  )

fig
```



# Summary Statistics by Group

Let's examine the differences between the two groups quantitatively:

```{r summary-stats}
data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    avg_age = mean(age),
    avg_education = mean(education),
    avg_income = mean(income),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 1,
    col.names = c("Sex", "N", "Avg Age", "Avg Education", "Avg Income"),
    caption = "Summary Statistics by Group"
  )
```

# Key Insights

From our parallel coordinate plots, we can observe:

1. **Two Distinct Patterns**: Even without labeling sex, two clear groups emerge
2. **Age-Income Relationship**: Higher age correlates with higher income (blue/men group)
3.  **Group Characteristics**:
   - **Group 1** (Blue/Men): Older, moderate education, higher income
   - **Group 2** (Red/Women): Younger, higher education, lower income

# Conclusion

Parallel Coordinate Plots are excellent for:

- Visualizing multivariate relationships
- Identifying patterns and clusters
- Exploring high-dimensional data
- Interactive data exploration

The key advantage is the ability to see **multiple dimensions** simultaneously, making hidden patterns visible!

---



```{r}
knitr::knit_exit()
```




