---
title: "Exercise with Parallel Coordiante Plots"
author: "Christophe Bontemps (SIAP)"
subtitle: '*An example*'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
  )
```

# Introduction

Parallel Coordinate Plots (PCP) are a powerful visualization technique for exploring multivariate data. They allow us to visualize multiple dimensions simultaneously by representing each variable as a vertical axis, with individual observations shown as connected lines across these axes.



# Load Libraries

```{r load-libraries}
library(tidyverse)
library(GGally)      # For ggparcoord
library(plotly)      # For interactive plots
library(viridis)     # For color palettes
```

# Data 

```{r generate-data, include=FALSE}
set.seed(123)

# Create dataset with CLEAR group separation
data <- tibble(
  id = 1:10,
  sex = c(rep("Men", 6), rep("Women", 4)),
  
  # Age: Men older (45-55), Women younger (25-35)
  age = c(
    48, 52, 45, 50, 47, 54,  # Men: 45-55 range
    28, 32, 26, 30           # Women: 25-35 range
  ),
  
  # Education: Men lower (12-14 years), Women higher (16-18 years)
  education = c(
    12, 14, 13, 12, 14, 13,  # Men: high school to some college
    18, 17, 18, 16           # Women: college to grad school
  ),
  
  # Income: Men high (70k-90k), Women moderate (40k-55k)
  income = c(
    85000, 78000, 82000, 90000, 75000, 88000,  # Men: high income
    48000, 52000, 45000, 50000                 # Women: moderate income
  )
)

```



```{r export-csv, include=FALSE}
# Export data to CSV

Mydata<- data %>%
  select(-sex)

write_csv(data, "Data/pcp_tutorial_data.csv")
write_csv(Mydata, "Data/pcp_Exercise_data.csv")

```

## Read CSV

```{r}
# read the data
Mydata <- read.csv("Data/pcp_Exercise_data.csv")
```

```{r}

# Display the data
Mydata %>%
  knitr::kable(caption = "Dataset for exercise (10 Observations)")
```
```


# Basic Parallel Coordinate Plot


```{r basic-pcp}
# Prepare data for PCP (select only numeric variables)
data_pcp <- data %>%
  select(id, age, education, income)

# Basic PCP using ggparcoord
ggparcoord(
  data_pcp,
  columns = 2:4,  # age, education, income
  groupColumn = "id",
  scale = "uniminmax"
) +
  labs(
    title = "Parallel Coordinate Plot: Age, Education, and Income",
    subtitle = "10 observations showing two distinct patterns",
    x = NULL,
    y = "Scaled Values (0-1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50", size = 11)
  )
```

**Observation:** Notice how the lines form two distinct groups, even though we haven't colored them by sex!

# Enhanced PCP with Transparency

Adding transparency (alpha) helps reveal overlapping patterns and density.

```{r enhanced-pcp}
ggparcoord(
  data_pcp,
  columns = 2:4,
  groupColumn = "id",
  scale = "uniminmax",
  alphaLines = 0.6
) +
  geom_line(size = 1.2) +
  labs(
    title = "Enhanced Parallel Coordinate Plot",
    subtitle = "With transparency to show overlapping patterns",
    x = NULL,
    y = "Scaled Values"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    panel.grid.major.x = element_line(color = "grey80", size = 0.5),
    panel.grid.minor = element_blank()
  )
```

# PCP with Hidden Grouping

Now let's color the lines by sex (but not show the legend), making the two groups visually distinct.

```{r pcp-colored}
# Add sex back for coloring
data_with_sex <- data %>%
  select(id, sex, age, education, income)

ggparcoord(
  data_with_sex,
  columns = 3:5,  # age, education, income
  groupColumn = "sex",
  scale = "uniminmax",
  alphaLines = 0.7
) +
  scale_color_manual(values = c("Men" = "#3498DB", "Women" = "#E74C3C")) +
  labs(
    title = "Parallel Coordinate Plot: Two Distinct Groups",
    subtitle = "Blue and red lines show different patterns (sex not labeled)",
    x = NULL,
    y = "Scaled Values (0-1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.text = element_text(size = 11),
    panel.grid.major.x = element_line(color = "grey80", size = 0.5)
  )
```

**Key Patterns:**

- **Blue lines** (Men): Higher on age and income, moderate on education
- **Red lines** (Women): Lower on age and income, higher on education

# Beautiful PCP with Viridis Colors

Using a gradient color scheme for a more sophisticated look.

```{r pcp-viridis}
ggparcoord(
  data_with_sex,
  columns = 3:5,
  groupColumn = "sex",
  scale = "uniminmax",
  alphaLines = 0.8
) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    title = "Parallel Coordinate Plot with Viridis Colors",
    subtitle = "Exploring relationships between age, education, and income",
    x = "Variables",
    y = "Normalized Values"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey50", size = 11),
    axis.text.x = element_text(face = "bold", size = 12),
    panel.background = element_rect(fill = "#f8f9fa", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major.x = element_line(color = "grey70", size = 0.8),
    panel.grid.major.y = element_line(color = "grey90", size = 0.3)
  )
```

# Interactive Parallel Coordinate Plot

Finally, let's create an **interactive version** using `plotly` that allows hovering and filtering.

```{r interactive-pcp, fig.height=7}
# Prepare data for plotly
plot_data <- data %>%
  mutate(
    color_code = ifelse(sex == "Men", 0, 1),
    label = paste0(
      "ID: ", id, "<br>",
      "Sex: ", sex, "<br>",
      "Age: ", age, "<br>",
      "Education: ", education, " years<br>",
      "Income: $", format(income, big.mark = ",")
    )
  )

# Create interactive parallel coordinates plot
fig <- plot_ly(
  data = plot_data,
  type = 'parcoords',
  line = list(
    color = ~color_code,
    colorscale = list(
      c(0, '#3498DB'),  # Men: Blue
      c(1, '#E74C3C')   # Women: Red
    ),
    showscale = FALSE
  ),
  dimensions = list(
    list(
      label = 'Age',
      values = ~age,
      range = c(20, 60)
    ),
    list(
      label = 'Education (years)',
      values = ~education,
      range = c(10, 20)
    ),
    list(
      label = 'Income ($)',
      values = ~income,
      range = c(30000, 90000)
    )
  )
)

fig <- fig %>%
  layout(
    title = list(
      text = "Interactive Parallel Coordinate Plot<br><sub>Hover over lines to see details | Drag axes to filter</sub>",
      font = list(size = 16)
    ),
    margin = list(l = 100, r = 100, t = 100, b = 50),
    plot_bgcolor = '#f8f9fa',
    paper_bgcolor = 'white'
  )

fig
```

## Interactive Features

The interactive plot above allows you to:

1. **Hover** over lines to see individual observation details
2. **Click and drag** on any axis to filter the data
3. **Reorder axes** by dragging axis labels
4. **Zoom** into specific ranges

# Summary Statistics by Group

Let's examine the differences between the two groups quantitatively:

```{r summary-stats}
data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    avg_age = mean(age),
    avg_education = mean(education),
    avg_income = mean(income),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 1,
    col.names = c("Sex", "N", "Avg Age", "Avg Education", "Avg Income"),
    caption = "Summary Statistics by Group"
  )
```

# Key Insights

From our parallel coordinate plots, we can observe:

1. **Two Distinct Patterns**: Even without labeling sex, two clear groups emerge
2. **Age-Income Relationship**: Higher age correlates with higher income (blue/men group)
3.  **Group Characteristics**:
   - **Group 1** (Blue/Men): Older, moderate education, higher income
   - **Group 2** (Red/Women): Younger, higher education, lower income

# Conclusion

Parallel Coordinate Plots are excellent for:

- Visualizing multivariate relationships
- Identifying patterns and clusters
- Exploring high-dimensional data
- Interactive data exploration

The key advantage is the ability to see **multiple dimensions** simultaneously, making hidden patterns visible!

---



```{r}
knitr::knit_exit()
```




