---
title: "International Merchandise Trade Statistics Highlights"
author: "Christophe"
output:
  html_document:
    toc: no
    code_folding: show
    df_print: paged
  pdf_document:
    toc: no
  word_document:
    toc: no
---

# Data analysis
We provide here some basic elements about the data set and descriptive statistics on some variables of interest.  

* First we load the data
* second we print a summary


```{r}
library(tidyverse)
```

```{r}
TradeData <- read.csv("https://www.unsiap.or.jp/on_line/2024/RAP/Data/TradeData.csv")
```

We loaded the data set and compile the  report on `r Sys.Date()`

## Summary statistics
Here are some summary statistics of the export value (*Stat_Value* variable), all products combined:

```{r }
summary(TradeData$Stat_Value)
```


```{r NbGoods}
#  This is the number of different goods 
TradeGoods <- sqrt(length(unique(TradeData$Principle_Export)))

```


The data set contains   **`r  nrow(TradeData) +1`  observations**, and  `r TradeGoods ` variables.

Here is a list of all variables available in the data set: 

```{r}
names(TradeData)
```


The variable *Principle_Export* list all the types of goods exported from Esdeegia. There are **`r  length(unique(TradeData$Principle_Export))`** different goods exported. Let's have the list of all these goods. 

```{r}
unique(TradeData$Principle_Export)

```



```{r}
p<- ggplot(TradeData) +
  aes(x = Stat_Value, fill = Principle_Imports) +
  geom_histogram(bins = 30L) +
  scale_fill_hue(direction = 2) +
  theme_minimal() +
coord_flip() +
  facet_wrap(vars(IMPORT_REGION))

p

```
```{r}
library(plotly)
ggplotly(p)
```

```{r}
knitr::knit_exit()
```

